Lemmas are general properties about predicates or functions. These properties
can be proved in isolation of the rest of the proof of a program by automatic or
(more often) interactive provers. Once this proof is done, the information that
it states can be safely used to simplify the reasoning in other, more complex
proofs, without having to prove it again. For example, if we state a lemma $L$
that says $P \Rightarrow Q$ in any case, if at some point in another proof, we
have to prove $Q$, and we know $P$, we can directly conclude by using the lemma
$L$ without having to perform again the reasoning that brings us from $P$ to
$Q$.


In the previous section, we said that recursive function can make proof harder
for SMT solvers. In such a case, lemmas can help us. We can write by ourselves
the proofs that require inductive reasoning for some properties that we state as
lemmas, and these lemmas can be used efficiently by SMT solvers to perform the
other proofs, related to our programs.


\levelThreeTitle{Syntax}


Again, we introduce lemmas using ACSL annotations. The syntax is
following:



\begin{CodeBlock}{c}
/*@
  lemma name_of_the_lemma { Label0, ..., LabelN }:
    property ;
*/
\end{CodeBlock}



This time, the properties we want to express do not depend on received
parameters (except for labels). So we express these properties for universally
quantified variables. For example, we can state this lemma, which is true, even
if it is trivial:



\begin{CodeBlock}{c}
/*@
  lemma lt_plus_lt:
    \forall integer i, j ; i < j ==> i+1 < j+1;
*/
\end{CodeBlock}



This proof can be performed using WP. The property is, of course, proved
using only Qed.



\levelThreeTitle{Example: properties of linear functions}


We can come back to our linear functions and express some interesting
properties about them:



\CodeBlockInput[12][22]{c}{linear-0.c}



For these proofs, Alt-ergo, will probably not be able to discharge all
generated verification conditions. In this case, CVC5 will certainly succeed
in finishing the remaining ones. We can then write the following example code:



\CodeBlockInput[24][52]{c}{linear-0.c}



If we do not give the lemmas provided earlier, Alt-ergo will not be able
to prove the proof that \CodeInline{fmin} is lower or equals to
\CodeInline{fmax}. With the lemmas it is however very easy for it since the
property is the simply an instance of the lemma
\CodeInline{ax\_monotonic\_pos}, the proof is then trivial as our lemma is
considered to be true when are not currently proving it. Note that on this
generalized version of \CodeInline{function}, CVC5 will be probably more
efficient to prove the absence of runtime errors.


\levelThreeTitle{Example: arrays and labels}


Later in this tutorial, we will see some kind of definitions for which it is
sometimes hard to reason about for SMT solvers when some mutations happen in
memory. Thus, we will often need lemmas to state relations about the content
of the memory between labels.


For now, let us illustrate with a first simple example. Consider the two following
predicates:


\CodeBlockInput[1][8]{c}{unchanged-sorted.c}


One could for example want to state that when an array is sorted, and some
mutations happen in memory (creating a new memory state), but the content of the
array remains unchanged, then the array is still sorted. This can be done with
the following lemma:


\CodeBlockInput[10][16]{c}{unchanged-sorted.c}


We state this lemma for two labels \CodeInline{L1} and \CodeInline{L2}, and
express that if any range in any array is sorted at \CodeInline{L1}, and
unchanged from \CodeInline{L1} to \CodeInline{L2}, then it is still sorted at
\CodeInline{L2}.


Note that this lemma is easily proved by SMT solvers. We will see later some
examples where it is not so easy to get a proof.


\levelThreeTitle{Check lemma}


Like assertions, it is possible to use a \CodeInline{check} version of lemmas:


\begin{CodeBlock}{c}
/*@
  check lemma name_of_the_lemma { Label0, ..., LabelN }:
    property ;
*/
\end{CodeBlock}


A \CodeInline{check lemma} annotation asks WP to generate a verification
condition. But, contrarily to standard lemmas, the knowledge that the
corresponding property is verified will not be added to the context of the
proofs. Thus, SMT solvers will not be able to use it for other proofs. This
kind of annotations can be useful to gain confidence about global annotations,
testing whether they can be used in some situations, or on the opposite, that
some unwanted properties cannot be proved.


There is no \CodeInline{admit lemma} annotation, these are \CodeInline{axiom}
annotations that are described in
Section~\ref{l2:acsl-logic-definitions-axiomatic}.


\levelThreeTitle{Exercises}



\levelFourTitle{Multiplication property}



Write a lemma that states that for three integers $x$, $y$ and $z$, if $x$ is
greater than or equal to $0$, if $z$ is greater than or equal to $y$, then
$x * z$ is greater than or equal to $x * y$.


This lemma will not be proved by SMT solvers. We provide a solution and the
corresponding Coq proof on the GitHub repository of this book.


\levelFourTitle{Locally sorted to globally sorted}
\label{l4:acsl-properties-lemmas-lsorted-gsorted}


The following program contains a function that requires an array to be sorted
in the sense that each element is lower than or equal to the element that
follows it and calls the binary search function.


\CodeBlockInput[66]{c}{ex-2-sorted-link.c}


Take back your proved binary search function from the
exercise~\ref{l4:acsl-properties-predicates-ex-bsearch}. You might notice that
the precondition of the binary search function is stronger than what we know in
precondition of the \CodeInline{bsearch\_callee} function. However, our precondition
implies that the array is globally sorted. Write a lemma that states that if
an array is \CodeInline{element\_level\_sorted} then it is \CodeInline{sorted}.
This lemma will probably not be proved by SMT solvers, all remaining properties
should be.

We provide a solution and the corresponding Coq proof on the GitHub repository
of this book.


\levelFourTitle{Sum of the N first integers}
\label{l4:acsl-properties-lemmas-n-first-ints}

Take back your solution to the
exercise~\ref{l4:acsl-properties-functions-n-first-ints} about the sum of the
N first integers. Write a lemma that states that the result of the call to the
logic function is $n*(n+1)/2$. This lemma will not be proved by SMT solvers.


We provide a solution and the corresponding Coq proof on the GitHub repository
of this book.


\levelFourTitle{Shift transitivity}
\label{l4:acsl-properties-lemmas-shift-trans}


The following program is composed of two functions. The first one is the
\CodeInline{shift\_array} function that shifts the elements of an array with a
given offset (named \CodeInline{shift}). The second performs two successive
shifts on the same array.


\CodeBlockInput[5]{c}{ex-4-shift-transitivity.c}


Complete the predicates \CodeInline{shifted} and \CodeInline{unchanged}.
The \CodeInline{shifted} predicate should use \CodeInline{shifted\_cell}.
The \CodeInline{unchanged} predicate should use \CodeInline{shifted}.
Complete the contract of the \CodeInline{shift\_array} function and prove
it using WP.


Express two lemmas about the \CodeInline{shifted} property.


The first one \CodeInline{shift\_ptr} should state that shifting a range
\CodeInline{fst+s1} to \CodeInline{last+s1} of an \CodeInline{array} with an
offset \CodeInline{s2} is equivalent to shifting the range \CodeInline{fst} to
\CodeInline{last} of the memory location \CodeInline{array+s1} with an offset
\CodeInline{s2}.



The second one should state that when the elements of an array are shifted a
first time with an offset \CodeInline{s1} and then a second time with an offset
\CodeInline{s2}, then the complete shift corresponds to a single shift with an
offset \CodeInline{s1+s2}.



The lemma \CodeInline{shift\_ptr} will not be proved by SMT solvers, we provide
a solution and the corresponding Coq proof on the GitHub repository of this
book. All remaining properties should be proved automatically.



\levelFourTitle{Shift sorted range}


The following program is composed of two functions. The function
\CodeInline{shift\_and\_search} shifts the element of an array and then
performs a binary search.


\CodeBlockInput[5]{c}{ex-5-shift-sorted.c}


Take back your proved binary search function from the
exercise~\ref{l4:acsl-properties-predicates-ex-bsearch}, modify the binary
search function, its contract and its proof in order to be able to search in
any range.


Use the \CodeInline{shift\_array} function proved in the previous exercise.


Complete the contract of the function \CodeInline{shift\_and\_search}. You might
notice that the precondition that requires the array to be sorted is not
validated, nor the postcondition of the caller. First, complete the lemma
\CodeInline{shifted\_still\_sorted} that should state that if a range is sorted
at some label and then shifted, the resulting range is still sorted, the
precondition should now be proved. Then, complete the lemma
\CodeInline{in\_array\_shifted} that should state that if a value is in a range
of values that is then shifted, it is still in the resulting new range. The
postcondition of the caller should now be proved.


The lemmas do not have to be proved by SMT solvers. We provide a solution and
the corresponding Coq proofs on the GitHub repository of this book.
